WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-28 articles/2014/Article-WSClean/wsclean.tex
Version as submitted to MWA collaborative review

\def\ANU{$^{1}$}
\def\CAASTRO{$^{2}$}
\def\Curtin{$^{3}$}
\def\UWisc{$^{4}$}
\def\MIT{$^{5}$}
\def\SKASA{$^{6}$}
\def\ASU{$^{7}$}
\def\Haystack{$^{8}$}
\def\RRI{$^{9}$}
\def\USydney{$^{10}$}
\def\CfA{$^{11}$}
\def\UW{$^{12}$}
\def\Victoria{$^{13}$}
\def\CASS{$^{14}$}
\def\Tata{$^{15}$}
\def\NRAO{$^{16}$}
\def\UMelbourne{$^{17}$}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
M.~E.~Bell\USydney$^,$\CAASTRO,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for widefield telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One such DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking \citep{awimager-2013}.

@@

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to successfully image any MWA observation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{The WSClean source code can be found at:\\\url{http://sourceforge.net/p/wsclean}}.

@@

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean. The pulsar (centre of image) displays 17\% circular polarization.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beam-formed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is somewhat confusing because they are not necessarily orthogonal.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. This requires four runs of the algorithm. Once these images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients and exoplanets. The pipeline includes WSClean to make the Stokes-parameter images.

@@

The available SKA computing power is estimated to be around $\sim$100 PetaFLOPS. Extrapolation of our results shows that the current imagers require $3$--$60$ petaFLOPS for SKA1 low alone. It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results show this is not the case when already using the $w$-stacking algorithm. Clearly it remains challenging to perform widefield imaging with acceptable performance, but some optimisations could be made for the SKA. For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging. Moreover, although longer baselines make the imaging more expensive, it is likely that it is not always required to image the full field of view when using the longest baselines.

>>>>

\def\ANU{$^{1}$}
\def\CAASTRO{$^{2}$}
\def\Curtin{$^{3}$}
\def\UWisc{$^{4}$}
\def\USydney{$^{5}$}
\def\MIT{$^{6}$}
\def\SKASA{$^{7}$}
\def\ASU{$^{8}$}
\def\Haystack{$^{9}$}
\def\RRI{$^{10}$}
\def\CfA{$^{11}$}
\def\UW{$^{12}$}
\def\Victoria{$^{13}$}
\def\CASS{$^{14}$}
\def\Tata{$^{15}$}
\def\NRAO{$^{16}$}
\def\UMelbourne{$^{17}$}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
M.~E.~Bell\USydney$^,$\CAASTRO,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
T.~Murphy\USydney$^,$\CAASTRO,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for wide-field telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One such DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking \citep{awimager-2013}.

@@

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to successfully image any MWA observation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The WSClean source code can be found at:\\\url{http://sourceforge.net/p/wsclean}}.

@@

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean. The pulsar (centre of image) displays 17\% circular polarization.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beam-formed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is somewhat confusing because they are not necessarily orthogonal.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. This requires four runs of the algorithm. Once these images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the MWA radio-sky monitor project (PI: Bell) that uses the MWA to search for transient and variable sources. The pipeline includes WSClean to make the Stokes-parameter images.

@@

The available SKA computing power is estimated to be around $\sim$100 PetaFLOPS. Extrapolation of our results shows that the current imagers require $3$--$60$ petaFLOPS for SKA1 low alone. It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results show this is not the case when already using the $w$-stacking algorithm. Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA. For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging. Moreover, although longer baselines make the imaging more expensive, it is likely that it is not always required to image the full field of view when using the longest baselines.
